---
title: "Analysis of impact of storms on population health and economy in USA"
output: 
  html_document:
    keep_md: true
---


```{r}
# setwd("~/Google Drive/GitHub/Reproducible-Research/Peer Assessment 2")
# download.file('https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2',destfile="data.csv")
data= read.csv("data.csv")
```

## Summary

Database is from U.S. National Oceanic and Atmospheric Administration's (NOAA) which tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and properties damage, covering period from the year 1950 to November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete. This report studies the number of fatalities and injuries, and economic damages of properties and crops valued in USD. It is found that several natural events led by tornado and flood is damaging population health and economy. Descriptions for variables can be found in original code book ([link](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)).


## Data Processing

Irrelevant variables will be removed in specific analysis conducted below. 

```{r}
# libraries
library(reshape2)
library(ggplot2)
require(gridExtra)
require(dplyr)

# new data set related to population health
ph=data[,c("EVTYPE","FATALITIES","INJURIES")]
phsum=as.data.frame(summarise(group_by(ph,EVTYPE),fatalities=sum(FATALITIES),injuries=sum(INJURIES)))
# remove events not threathening population health
phsumna=phsum[phsum$fatalities!=0 & phsum$injuries!=0,]


# new data set related to economic damage
ec=data[,c("EVTYPE","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")]
# apply damage explanation to damage amount
ec$PROPDMGEXP=as.character(ec$PROPDMGEXP)
ec$PROPDMGEXP=gsub("\\-|\\+|\\?|0","1",ec$PROPDMGEXP)
ec$PROPDMGEXP=gsub("b|B","9",ec$PROPDMGEXP)
ec$PROPDMGEXP=gsub("H|h","2",ec$PROPDMGEXP)
ec$PROPDMGEXP=gsub("k|K","3",ec$PROPDMGEXP)
ec$PROPDMGEXP=gsub("m|M","6",ec$PROPDMGEXP)
ec$PROPDMGEXP=as.numeric(ec$PROPDMGEXP)
ec$PROPDMGEXP[is.na(ec$PROPDMGEXP)]=0

ec$CROPDMGEXP=as.character(ec$CROPDMGEXP)
ec$CROPDMGEXP=gsub("\\-|\\+|\\?|0","1",ec$CROPDMGEXP)
ec$CROPDMGEXP=gsub("b|B","9",ec$CROPDMGEXP)
ec$CROPDMGEXP=gsub("H|h","2",ec$CROPDMGEXP)
ec$CROPDMGEXP=gsub("k|K","3",ec$CROPDMGEXP)
ec$CROPDMGEXP=gsub("m|M","6",ec$CROPDMGEXP)
ec$CROPDMGEXP=as.numeric(ec$CROPDMGEXP)
ec$CROPDMGEXP[is.na(ec$CROPDMGEXP)]=0

ec$PROPDMG = ec$PROPDMG*10^ec$PROPDMGEXP
ec$CROPDMG = ec$CROPDMG*10^ec$CROPDMGEXP

ecsum=as.data.frame(summarise(group_by(ec,EVTYPE),properties=sum(PROPDMG),crops=sum(CROPDMG)))
# remove events not causing economic damages
ecsumna=ecsum[ecsum$properties!=0 & ecsum$crops!=0,]
```

## Results

### Events that are most harmful to population health

Two variables in the original dataset are related to population health: 

+ Fatalities
+ Injuries    

The analysis will focus on these two variables to evaluate the scale of damage events have on population health.    
```{r}
phbytotal= phsumna %>%
  mutate(total=fatalities+injuries) %>%
  arrange(desc(total))
head(phbytotal,n=20)
```

By observing priliminary analysis, it can be concluded that Tornado is the only event that can significantly damage population health, causing **`r phbytotal$fatalities[1]`** fatalities and **`r round(phbytotal$injuries[1]/10^3,1) `** thousands injuries.   
In total **`r nrow(phsumna) `** events have record of injuries or fatalities. To generate the most relevant results, only events that have caused more than 1000 injuries or injuries and fatalities combined, or caused more than 100 fatalities will be further analysed. 

The analysis will choose three numbers - fatalities, injuries, and fatalities&injuries combined - to suggest the top nature events harmful to the population. 

#### Rank by combination of fatalities & injuries
```{r}
# evaluate overall damage
phbytotal= filter(phbytotal,total>= 1000)

phmelt= melt(phbytotal[-1,],measure.vars=c("fatalities","injuries"),id.vars = c("EVTYPE"))
phmelt$EVTYPE= factor(phmelt$EVTYPE,levels=arrange(phbytotal,total)$EVTYPE)

ggplot(phmelt,aes(y=value,x=EVTYPE,fill=variable))+geom_bar(stat = "identity")+labs(x="Event Type", y="Number of fatalities/injuries per event")+ coord_flip()+geom_text(aes(label = value), size = 3)
```

Besides Tornado, these events listed are the top natural disasters that affect population health. 


#### Rank by fatalities and injuries separately
```{r}
phbyfatal=
  phsumna %>%
  arrange(desc(fatalities)) %>%
  filter(fatalities>= 100) %>%
  select(-injuries)

phbyfatal$EVTYPE= factor(phbyfatal$EVTYPE,levels=arrange(phbyfatal,fatalities)$EVTYPE)
p1= ggplot(phbyfatal[-1,],aes(y=fatalities,x=EVTYPE))+geom_bar(stat = "identity",fill="indianred2")+labs(x="Event Type", y="Number of fatalities\n per event")+ coord_flip()+geom_text(aes(label = fatalities), size = 3.5)


# only look at events with injuries over 1000
phbyinj= 
  phsumna %>%
  arrange(desc(injuries)) %>%
  filter(injuries>= 1000) %>%
  select(-fatalities)

phbyinj$EVTYPE= factor(phbyinj$EVTYPE,levels=arrange(phbyinj,injuries)$EVTYPE)
p2= ggplot(phbyinj[-1,],aes(y=injuries,x=EVTYPE))+geom_bar(stat = "identity",fill="turquoise3")+labs(x="",y="Number of injuries\n per event")+ coord_flip()+geom_text(aes(label = injuries), size = 3)

grid.arrange(p1, p2, ncol = 2)
```

Given the ranking of severity, the authority should be aware of these natural disasters accordingly and create procedures in pre-warning, dismissing and rescuing residents to reduce injuries and fatalities. 

### Events that create greatest most harmful economic consequences

Two variables in the original dataset are related to economic damages: 

+ properties damages
+ crops damages    

The analysis will focus on these two variables to evaluate the scale of damage events have on economy.    

```{r}
ecbytotal= ecsumna %>%
  mutate(total=properties+crops) %>%
  arrange(desc(total))
head(ecbytotal,n=20)
```

By observing priliminary analysis, it can be concluded that Flood is the only event that can significantly harm economy by causing $**`r round(ecbytotal$properties[1]/10^9,1) `**bn USD properties damages and $**`r round(ecbytotal$crops[1]/10^9,1)`**bn USD crops damages.   
In total **`r nrow(ecsumna) `** events have record of economic damages. To generate the most relevant results, only events that have caused more than $10bn damages on properties or properties & crops combined, or damages more than $1bn on crops will be further analysed. 

The analysis will choose three quantified amount of damages on properties, crops, and properties&crops combined to suggest the top nature events harmful to the economy. 

#### Rank by combination of properties & crops 
```{r}
# evaluate overall damage
ecbytotal= filter(ecbytotal,total>= 10^10)

ecmelt= melt(ecbytotal[-1,],measure.vars=c("properties","crops"),id.vars = c("EVTYPE"))
ecmelt$EVTYPE= factor(ecmelt$EVTYPE,levels=arrange(ecbytotal,total)$EVTYPE)


ggplot(ecmelt,aes(y=value,x=EVTYPE,fill=variable))+geom_bar(stat = "identity")+labs(x="Event Type", y="Number of properties/crops damages per event")+ geom_text(aes(label = round(value/10^9,2)), size = 3) +scale_y_continuous(breaks=c(seq(10^9,10^11,10^10))) + coord_flip() 

```

Besides Flood, these events listed are the top natural disasters that do economic damages. 

#### Rank by properties and crops separately

```{r}
# only look at events with properties over $10bn
ecbyprop=
  ecsumna %>%
  arrange(desc(properties)) %>%
  filter(properties>= 10^10) %>%
  select(-crops)

ecbyprop$EVTYPE= factor(ecbyprop$EVTYPE,levels=arrange(ecbyprop,properties)$EVTYPE)
p1= ggplot(ecbyprop[-1,],aes(y=properties,x=EVTYPE))+geom_bar(stat = "identity",fill="indianred2")+labs(x="Event Type", y="Properties damages\n per event ($bn)")+ coord_flip()+geom_text(aes(label = round(properties/10^9,1)), size = 3)


# only look at events with crops over $10bn
ecbycrop=
  ecsumna %>%
  arrange(desc(crops)) %>%
  filter(crops>= 10^9) %>%
  select(-properties)

ecbycrop$EVTYPE= factor(ecbycrop$EVTYPE,levels=arrange(ecbycrop,crops)$EVTYPE)
p2= ggplot(ecbycrop[-1,],aes(y=crops,x=EVTYPE))+geom_bar(stat = "identity",fill="turquoise3")+labs(x="",y="Crops damages\n per event ($bn)")+ coord_flip()+geom_text(aes(label = round(crops/10^9,1)), size = 3.5)

grid.arrange(p1, p2, ncol = 2)

```

Given the ranking of severity, the authority should be aware of these natural disasters accordingly and create procedures to reduce economic damages.


### Limitations 

There are several limitations besides those listed in original codebook.  	

+ All time period covered by original database is used in analysis. As given in the original codebook, it was only until recent years that such natural disasters were properly recorded. Including all time period may have an impact on the results since it is probable that only serious events were recorded in the past, casting the total quantified damages caused. 	
+ Some names of event may be spelled differently or misspelled. Events may be overlapped since multiple events may have happened at the same time. This will separate data within same classification and lead to inaccuracy. The problem is not addressed in the data processing due to tight deadline.  
+ Economic damages may be estimated figures. This will surely affect the accuracy of results.		
